Evolving Dynamics: The U.S. Labor Force from 1972 to 2022

### 2024-03-18 paper/paper.qmd
updated references

Over the past fifty years, the U.S. labor force has undergone significant transformations, reflecting broader socio-economic shifts, technological advancements, and changing demographic patterns. These changes have had profound implications for employment practices, workplace cultures, and the economic opportunities available to different segments of the population. However, despite these overall shifts, women in the workforce have continued to face enduring challenges, including gender pay gaps, underrepresentation in leadership positions, and the persistent juggling act of work-family balance. This paper examines the evolving dynamics of the U.S. labor force from 1972 to 2022, focusing on how these broader changes have intersected with the experiences of women in the workplace.

As we focus on the exploration of labor force trends with an examination of the specific, ongoing challenges that women encounter in the workplace, we will touch on how our investigation seeks to bridge the gap on general labor trends and women's workplace challenges by employing data from the U.S. General Social Survey (GSS). This allows us to dissect how five decades of labor market evolution have sculpted the professional landscape for women. We aim to uncover the intricate ways in which shifts in the labor force have impacted women's experiences in the workforce.

The U.S. General Social Survey (GSS) has been a pivotal source of data, employing a sampling approach to ensure representative insights into the American public's attitudes, behaviors, and attributes since 1972. This methodology involves selecting participants to mirror the U.S. population's demographic makeup, including variations by region, age, gender, and socio-economic status. Notably, the GSS adapted its data collection methods in response to the COVID-19 pandemic, transitioning to online and phone surveys to maintain participant safety and response rates. These adjustments, while necessary, introduced potential implications for trend analysis, including variations in response rates and the potential for digital access disparities. Despite these challenges, our analysis of the GSS data reveals striking findings. For instance, from 1972 to 2022, women's participation in the workforce increased significantly. Initially, the workforce was predominantly male, with 570 men working full time compared to 191 women in 1972 but by 2022, the number of women working full-time had increased to 728 from 191 in 1972, showing substantial progress towards gender equality in the workplace. However, the persistence of work-family balance challenges is evident, underscoring the ongoing struggle to integrate professional and personal responsibilities. This paper will reveal that despite an increase in participation rates and educational attainment among women, numerous deep-rooted barriers persist. Crucially, this paper highlights the diversity of women's experiences in the labor market, underscoring the importance of an intersectional approach to understanding workplace challenges.

@@

Data used in this paper are retrieved from the U.S. General Social Survey (GSS).
The General Social Survey (GSS) is an ongoing interview survey of U.S. households conducted by the National Opinion Research Center. It has provided a wealth of data on contemporary American society for approximately 35 years by measuring social change and trends and constants in the adult population's attitudes, behaviors, and attributes. It includes a questionnaire such as demographics, attitudes, and ideas towards certain beliefs, and even extends economic ideologies such as spending priorities. The goal is to have high-quality data available to social scientists and researchers. Over the years, the GSS has worked continuously to make the survey better. One major improvement worthy of mentioning is the addition of a Spanish version, since there is a big demographic of the U.S. that speaks Spanish, this would provide a significant increase in the coverage of U.S. household responses.

@@

The COVID-19 pandemic posed unprecedented challenges to data collection efforts, including those of the General Social Survey (GSS), necessitating significant adjustments in methodology that may impact the reliability and comparability of the data collected during this period. In response to social distancing guidelines, the GSS likely had to transition from traditional in-person interviews to remote data collection methods, such as online surveys or telephone interviews. While essential for public health, these shifts could introduce biases, particularly affecting response rates and the representativeness of certain demographic groups less comfortable or accessible via these technologies. The pandemic's uneven impact across various regions, economic statuses, and racial demographics further complicates sampling efforts, potentially skewing the sample composition.

@@

R [@citeR] was the language and environment used for this paper as well as throughout the data cleaning process, with different packages such as tidyverse [@tidyverse], ggplot2 [@ggplot2], dplyr [@dplyr], tidyr [@tidyr], knitr [@knitr] packages.

@@

The trend of women participating in full-time work and education is also increasing. Upon taking a look at @fig-eduTojob, the number of graduate, undergraduate, and full-time lines is trending upwards. There are some significant jumps in the lines, however, this could be due to the data collection amount as these jumps are also observed in some other figures such as figure 8. With the attitudes steadily becoming more positive towards women in the workplace, the pursuit of higher education and full-time employment has also significantly increased. Something interesting observed is that the peak number of full-time workers is less than the most recent data. To gain insight into why this may be, additional research was done. In the Forbes article, it talks about the significant increase of women in higher education. With full-time working women decreasing, it could be attributed to them obtaining higher education instead of choosing to work now. Overall, this is a good trend as more women are pursuing higher education, and should allow them to obtain more skilled jobs in the future.

@@

Financial crises always have a devastating impact on society, how does it affect the women in the workforce? The financial crisis this section refers to is the 2008 housing bubble. There are also a few things that the data captured in the results section. In @fig-education, there is a sharp decrease in every line in the chart. The number of women that are participating in Full-time work, studying for graduation, or bachelor's all decreased. Upon doing additional research, in this paper. When an economic crisis occurs, the gender norms and biases we often forget about resurface. During the crisis, companies often cut costs by reducing labor. As a result, companies often decide to start with female workers because they are often seen as “secondary income” for the family. Something the article mentioned is loans taken out by women were impacted more, as lenders became more skeptical which leads to an impact on women-owned businesses. The number of graduate and bachelor women students also decreased, this could be because that family can no longer support the education, and decided that this would be a method of cutting living expenses. One example the paper mentioned is that women in India are expected to eat less as a result of reduced income to ensure that the rest of the family has enough food. As a result, women are often affected more by economic crises.

@@

With the data from the U.S. General Social Survey, the paper utilized it to gain further insight into the challenges women are enduring in the workplace. The time frame is slightly different in each figure but mostly focuses from around 1970 to 2022. To gain further insight, we looked at questions that could gain insight into attitudes towards women working and questions that could be affecting factors to women participating in the workforce. The data revealed that women overall have gained significant progress, but there are still many factors that need to be taken into consideration. Progress is indicated by a steady increase in positive attitudes towards women in the workplace, and also an increasing amount of women pursuing higher education. While these are good indicators of progress, there is still a lot to be uncovered. These questions do not provide for direct indication of the gender pay gap which is often the most brought up issue. More research and data gathering would need to be done to provide further insight into this tip-of-the-iceberg problem. We must continue to work to provide gender equality in the workplace, where there is still a lot of room for everyone to contribute to a more equal working environment. Let’s continue the momentum and ensure we continue working for an equal workforce.

>>>>

Over the past fifty years, the U.S. labor force has undergone significant transformations, reflecting broader socio-economic shifts, technological advancements, and changing demographic patterns. These changes have had profound implications for employment practices, workplace cultures, and the economic opportunities available to different segments of the population. However, despite these overall shifts, women in the workforce have continued to face enduring challenges, including gender pay gaps, underrepresentation in leadership positions, and the persistent juggling act of work-family balance [@leadership_women]. This paper examines the evolving dynamics of the U.S. labor force from 1972 to 2022, focusing on how these broader changes have intersected with the experiences of women in the workplace.

As we focus on the exploration of labor force trends with an examination of the specific, ongoing challenges that women encounter in the workplace, we will touch on how our investigation seeks to bridge the gap on general labor trends and women's workplace challenges by employing data from the U.S. General Social Survey (GSS) [@GSS]. This allows us to dissect how five decades of labor market evolution have sculpted the professional landscape for women. We aim to uncover the intricate ways in which shifts in the labor force have impacted women's experiences in the workforce.

The U.S. General Social Survey (GSS) [@GSS] has been a pivotal source of data, employing a sampling approach to ensure representative insights into the American public's attitudes, behaviors, and attributes since 1972. This methodology involves selecting participants to mirror the U.S. population's demographic makeup, including variations by region, age, gender, and socio-economic status. Notably, the GSS adapted its data collection methods in response to the COVID-19 pandemic, transitioning to online and phone surveys to maintain participant safety and response rates. These adjustments, while necessary, introduced potential implications for trend analysis, including variations in response rates and the potential for digital access disparities. Despite these challenges, our analysis of the GSS data reveals striking findings. For instance, from 1972 to 2022, women's participation in the workforce increased significantly. Initially, the workforce was predominantly male, with 570 men working full time compared to 191 women in 1972 but by 2022, the number of women working full-time had increased to 728 from 191 in 1972, showing substantial progress towards gender equality in the workplace. However, the persistence of work-family balance challenges is evident, underscoring the ongoing struggle to integrate professional and personal responsibilities. This paper will reveal that despite an increase in participation rates and educational attainment among women, numerous deep-rooted barriers persist. Crucially, this paper highlights the diversity of women's experiences in the labor market, underscoring the importance of an intersectional approach to understanding workplace challenges.

@@

Data used in this paper are retrieved from the U.S. General Social Survey (GSS) [@GSS].
The General Social Survey (GSS) is an ongoing interview survey of U.S. households conducted by the National Opinion Research Center. It has provided a wealth of data on contemporary American society for approximately 35 years by measuring social change and trends and constants in the adult population's attitudes, behaviors, and attributes. It includes a questionnaire such as demographics, attitudes, and ideas towards certain beliefs, and even extends economic ideologies such as spending priorities. The goal is to have high-quality data available to social scientists and researchers. Over the years, the GSS has worked continuously to make the survey better. One major improvement worthy of mentioning is the addition of a Spanish version, since there is a big demographic of the U.S. that speaks Spanish, this would provide a significant increase in the coverage of U.S. household responses.

@@

The COVID-19 pandemic posed unprecedented challenges to data collection efforts, including those of the General Social Survey (GSS) [@GSS], necessitating significant adjustments in methodology that may impact the reliability and comparability of the data collected during this period. In response to social distancing guidelines, the GSS likely had to transition from traditional in-person interviews to remote data collection methods, such as online surveys or telephone interviews. While essential for public health, these shifts could introduce biases, particularly affecting response rates and the representativeness of certain demographic groups less comfortable or accessible via these technologies. The pandemic's uneven impact across various regions, economic statuses, and racial demographics further complicates sampling efforts, potentially skewing the sample composition.

@@

R [@citeR] was the language and environment used for this paper as well as throughout the data cleaning process, with different packages such as tidyverse [@tidyverse], ggplot2 [@ggplot2], dplyr [@dplyr], tidyr [@tidyr], knitr [@knitr], janitor [@janitor], lubridate [@lubridate] packages.

@@

The trend of women participating in full-time work and education is also increasing. Upon taking a look at @fig-eduTojob, the number of graduate, undergraduate, and full-time lines is trending upwards. There are some significant jumps in the lines, however, this could be due to the data collection amount as these jumps are also observed in some other figures such as figure 8. With the attitudes steadily becoming more positive towards women in the workplace, the pursuit of higher education and full-time employment has also significantly increased. Something interesting observed is that the peak number of full-time workers is less than the most recent data. To gain insight into why this may be, additional research was done. In the Forbes article, it talks about the significant increase of women in higher education. With full-time working women decreasing, it could be attributed to them obtaining higher education instead of choosing to work now. Overall, this is a good trend as more women are pursuing higher education, and should allow them to obtain more skilled jobs in the future. [@educ_women]

@@

Financial crises always have a devastating impact on society, how does it affect the women in the workforce? The financial crisis this section refers to is the 2008 housing bubble. There are also a few things that the data captured in the results section. In @fig-education, there is a sharp decrease in every line in the chart. The number of women that are participating in Full-time work, studying for graduation, or bachelor's all decreased. Upon doing additional research, in this paper [@impact_women], when an economic crisis occurs, the gender norms and biases we often forget about resurface. During the crisis, companies often cut costs by reducing labor. As a result, companies often decide to start with female workers because they are often seen as “secondary income” for the family. Something the article mentioned is loans taken out by women were impacted more, as lenders became more skeptical which leads to an impact on women-owned businesses. The number of graduate and bachelor women students also decreased, this could be because that family can no longer support the education, and decided that this would be a method of cutting living expenses. One example the paper mentioned is that women in India are expected to eat less as a result of reduced income to ensure that the rest of the family has enough food. As a result, women are often affected more by economic crises.

@@

With the data from the U.S. General Social Survey [@GSS], the paper utilized it to gain further insight into the challenges women are enduring in the workplace. The time frame is slightly different in each figure but mostly focuses from around 1970 to 2022. To gain further insight, we looked at questions that could gain insight into attitudes towards women working and questions that could be affecting factors to women participating in the workforce. The data revealed that women overall have gained significant progress, but there are still many factors that need to be taken into consideration. Progress is indicated by a steady increase in positive attitudes towards women in the workplace, and also an increasing amount of women pursuing higher education. While these are good indicators of progress, there is still a lot to be uncovered. These questions do not provide for direct indication of the gender pay gap which is often the most brought up issue. More research and data gathering would need to be done to provide further insight into this tip-of-the-iceberg problem. We must continue to work to provide gender equality in the workplace, where there is still a lot of room for everyone to contribute to a more equal working environment. Let’s continue the momentum and ensure we continue working for an equal workforce.
